Think deep for reef resilience

### 2025-08-16 paper/paper_draft.qmd
Simplifying notation
Streamlining time series clustering description
Removing some unnecessary detail

Implementation details, e.g., about size and shape of matrices, are not necessary for this kind of paper. If needed, they can be explained in supplementaries.

Add section in supplementary for expanded info on time series clustering (may remove, as it's described in cited paper...).

Also some comments for @eco-ben

$$
O:I = \frac{(S_O/N_O)}{(S_I/N_I)}
$$ {#eq-so-to-si}

@@

To identify groups of reefs that are resilient, intermediate, and non-resilient, we
clustered reefs into three clusters based on their temporal behaviour as indicated by coral
cover over time, using timeseries from 2025-2075. The Complexity-Invariant Distance (CID)
between each timeseries was calculated to identify timeseries that behave similarly
[@batista2014]. The CID corrects the Euclidean distance between pairs of timeseries (which
accounts for magnitude differences) with the complexity of each timeseries (obtained through
a measure of their variability) to ensure reefs are clustered based on their underlying
temporal dynamics rather than simply their average coral cover. The process creates a
distance matrix of n \* n dimensions (where n is the number of timeseries in the group) that
is clustered using a k-medoids clustering algorithm [cf. @steinmann2020]. The three reef
clusters are assigned a qualitative level from "low", "medium" and "high" based on their
relative median timeseries over the 2030-2060 period. Assigning cluster labels based on this
period prevents cluster labels being assigned during the steeply declining period from
2025-2030 and during the post-2060 period when many reefs have cover below 5% of their area.

Timeseries clustering analyses were performed at the management area and reefal bioregion
scale (defined in GBRMPA [-@GBRMPA2024]). Additional analyses were also performed at the
whole-of-GBR scale. These were found to lead to identical conclusions and so are not shown
here. To compare timeseries clustering results across GCMs, management area scale clustering
results are used to preserve coarse scale latitudinal differences between regions, while
maintaining ease of comparison and visualisation. These comparisons include quantification
of differences between low and high cover clusters across management regions under each GCM
using the Complexity-Invariant Distance between cluster median timeseries. To assess how
different reef characteristics influence timeseries clustering at a finer scale, bioregion
results are used to capture patterns occurring within individual clusters of reefs. Results
of all analyses at all clustering scales can be found at Grier and Iwanaga [-@grier2025] for
those interested.

Reefs in each timeseries cluster were compared to identify characteristics that differ
between low, medium and high coral cover clusters. These characteristics include reef median
depth, mean timeseries DHW, mean timeseries DHW tolerance level, Pearson’s correlation
between total coral cover and inversed DHW timeseries, proportion of coral habitable area
occupied by corals in the first timestep, ratio of source-to-sink larval connectivity and
total strength of all larval connections between reefs. To analyse the DHW tolerance levels
of reefs, heritable DHW tolerance values were tracked over timeseries in a matrix of shape
timestep \* taxa \* reef \* scenario matrix.

@@

>>>>

$$
O:I = \frac{S_O/N_O}{S_I/N_I}
$$ {#eq-so-to-si}

@@

Reefs were clustered into three qualitative categories based on their temporal behaviour
indicated by coral cover over time. The three categories were Low, Medium, and High,
indicating their relative ability to maintain coral cover and thus their structural
resilience. The Complexity-Invariant Distance (CID) between each time series was calculated
to identify similarities in temporal behaviour [@batista2014]. The CID applies a correction
factor to the Euclidean distance between pairs of time series to obtain a measure of their
similarity or dissimilarity. The adopted approach ensures reefs are clustered based on their
temporal dynamics as opposed to, for example, naively clustering by their mean coral cover.
Clustering similar time series is achieved through the use of k-medoids clustering [cf.
@steinmann2020]. Further details on the time series clustering approach are found in the
supplementary materials.

The time series clustering was applied over the time frame between 2030 and 2060. Spatially,
the analyses were conducted for each GBR management area and at the scale of each reefal
bioregion (defined in GBRMPA [-@GBRMPA2024]). Analyses were also performed at the
whole-of-GBR scale but were found to lead to identical results **\[TI: Ben, compared to
what, the bioregion or the management area results?\]** and so are not shown here. To
compare time series clustering results across GCMs, management area scale clustering results
are used to preserve coarse scale latitudinal differences between regions, while maintaining
ease of comparison and visualisation. These comparisons include quantification of
differences between low and high cover clusters across management regions under each GCM
using the CID between cluster median time series. To assess how different reef
characteristics influence time series clustering at a finer spatial scale, bioregion results
are used to capture patterns occurring within individual clusters of reefs. Results of all
analyses at all clustering scales can be found at Grier and Iwanaga [-@grier2025] for those
interested.

Reefs in each time series cluster were compared to identify characteristics that differ
between Low, Medium and High categories. These characteristics include reef median depth,
mean time series DHW, mean time series DHW tolerance level **\[TI: Ben, still correct? Still
included somewhere?\]**, Pearson’s correlation between total coral cover and inverse DHW
time series, proportion of coral habitable area occupied by corals in the first timestep,
ratio of source-to-sink larval connectivity and total strength of all larval connections
between reefs.

@@

**\[TI: To be revisited\]**

The CID corrects the Euclidean distance between pairs of time series (which accounts for
magnitude differences) with the complexity of each time series (obtained through a measure
of their variability) to ensure reefs are clustered based on their underlying temporal
dynamics rather than simply their average coral cover.

The process creates a distance matrix of n \* n dimensions (where n is the number of time
series in the group) that is clustered using a k-medoids clustering algorithm [cf.
@steinmann2020]. The three reef clusters are assigned a qualitative categories from Low,
Medium and High.
